Lossy compression of DNA sequencing quality data

### 2015-04-15 manuscript/q-compression.tex
Various changes in wording and formatting.

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- QualComp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  Bzip2, with the black asterisk corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure.  The lossless compression of each data set using BZip2
ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
asterisks on the abscissa. The compression methods tend to cluster
together across the different data sets. Across all data sets, the
0-degree polynomial regression, profile encodings, and QualComp have
the lowest bits/base-call.

QualComp with the rate parameter set to 100 bits/read has the lowest
\textsc{mse}, but requires 10-15x more storage than the profile
encoding methods for only a 2-3x reduction in \textsc{mse}.  When
QualComp's rate parameter is set to match the profile encoding
methods, QualComp performs slightly worse in terms of \textsc{mse}.
In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
QualComp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of
17.29.  Using 256-profile encoding requires less storage (0.079
bits/bp) and has a lower \textsc{mse} (11.85).

@@

The majority of compression methods retain more base-pairs after
preprocessing than the uncompressed sequences (Figure
\ref{fig:preprocessing}).  In general, as a given compression model
increases in complexity, i.e., as the number of profiles, polynomial
degrees, or rate increases, the amount of base-pairs kept approaches
the number base-pairs kept using the uncompressed sequences.  The
compression methods on the \textit{Mus Musculus} data set have the
greatest proportion of retained base-pairs compared to the
uncompressed sequences.  The \textit{Escherichia coli} str. K-12
MG1655 MiSeq data set has the smallest range.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
outperforms all others in all metrics (Table
\ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
%using the uncompressed reads outperforms all compression methods in
%terms of \textsc{lap}, NG50, relocations, translations, and
%inversions.
Among the compression methods, the 256-, 64-, 128-profile encoding had
the highest ranks, followed by QualComp with rates 10 bits/read, 30
bits/read, 100 bits/read, then 7th-degree polynomial regression,
followed by the QualComp with rate 6 bits/read, the 2-bin encoding,
and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.

@@

Among the different QualComp rate parameters, the 10 bits/read rate
ranked highest overall, outperforming the other rate parameters in
terms of corrected N50, least missing reference bases, \textsc{snp}s,
and indels $>$5bp.  With the exception of the 6 bits/read rate, the
assemblies decrease in rank with the increase in the rate parameter in
terms of corrected N50, and least missing reference bases. This trend
runs counter to the the decrease in \textsc{mse} of the different
rates.

\begin{figure}[!tbp]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}.
  Rankings above the median value are in cyan, those below in
  magenta}
  \label{fig:assembly_ranks}
\end{figure}

@@

The best compression method in terms of similarity with the
uncompressed reads is QualComp with rate 100 bits/read, followed by
QualComp with rate 30 bits/read, 256-, and 128-profile encoding, then
2-bin encoding, 64-profile encoding, QualComp with rate 10 bits/read,
7th-degree polynomial regression, QualComp with rate 6 bits/read, and
finally, 5th-degree through 0-degree polynomial regression.

@@

Most of the compression methods did not vary greatly in terms of the
number of reads that were mapped \emph{only} by the compression
method; however, there is a sizable difference in the amount of reads
that are originally mapped, but unmapped by the compressed methods.
QualComp with rate 100 bits/read results in the fewest missing
original read alignments (159).  Increasing the regression model
polynomial degree results in a decreasing amount of reads that are
originally mapped, but unmapped by the regression model (40,931 to
1,134 reads for 0-degree and 7th-degree, respectively).  There is no
such trend for reads that are mapped only by the regression model.

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the number
  of reads that are kept or discarded after applying compression with
  respect to the original, uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
  \cline{2-12}
& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
\cline{2-12}
 & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
\cline{2-12}
& proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
 & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
\cline{2-12}
& proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
\end{tabular}
\end{table*}

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
\textsc{QualComp}~\citep{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

@@

\section*{Acknowledgement}
We thank the organizers and participants of the 2014 Bioinformatics
Exchange for Students and Teachers (BEST) Summer School, which was
funded by the offices of the Dean of The Graduate School at University
of Maryland and the Rektor of University of T\"{u}bingen.

>>>>

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure.  The lossless compression of each data set using \textsc{bz}ip2
ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
asterisks on the abscissa. The compression methods tend to cluster
together across the different data sets. Across all data sets, the
0-degree polynomial regression, profile encodings, and \textsc{q}ual\textsc{c}omp have
the lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100 bits/read has the lowest
\textsc{mse}, but requires 10-15x more storage than the profile
encoding methods for only a 2-3x reduction in \textsc{mse}.  When
\textsc{q}ual\textsc{c}omp's rate parameter is set to match the profile encoding
methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms of \textsc{mse}.
In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of
17.29.  Using 256-profile encoding requires less storage (0.079
bits/bp) and has a lower \textsc{mse} (11.85).

@@

The majority of compression methods retain more base-pairs after
preprocessing than the uncompressed sequences (Figure
\ref{fig:preprocessing}).  In general, as a given compression model
increases in complexity, i.e., as the number of profiles, polynomial
degrees, or rate increases, the amount of base-pairs kept approaches
the number base-pairs kept using the uncompressed sequences.  The
compression methods on the \textit{Mus musculus} data set have the
greatest proportion of retained base-pairs compared to the
uncompressed sequences.  The \textit{Escherichia coli} str. K-12
MG1655 MiSeq data set has the smallest range.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
outperforms all others in all metrics (Table
\ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
%using the uncompressed reads outperforms all compression methods in
%terms of \textsc{lap}, NG50, relocations, translations, and
%inversions.
Among the compression methods, the 256-, 64-, 128-profile encoding had
the highest ranks, followed by \textsc{q}ual\textsc{c}omp with rates 10 bits/read, 30
bits/read, 100 bits/read, then 7th-degree polynomial regression,
followed by the \textsc{q}ual\textsc{c}omp with rate 6 bits/read, the 2-bin encoding,
and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.

@@

Among the different \textsc{q}ual\textsc{c}omp rate parameters, the 10 bits/read rate
ranked highest overall, outperforming the other rate parameters in
terms of corrected N50, least missing reference bases, \textsc{snp}s,
and indels $>$5bp.  With the exception of the 6 bits/read rate, the
assemblies decrease in rank with the increase in the rate parameter in
terms of corrected N50, and least missing reference bases. This trend
runs counter to the the decrease in \textsc{mse} of the different
rates.

\begin{figure}[!tbp]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}

@@

The best compression method in terms of similarity with the
uncompressed reads is \textsc{q}ual\textsc{c}omp with rate 100 bits/read, followed by
\textsc{q}ual\textsc{c}omp with rate 30 bits/read, 256-, and 128-profile encoding, then
2-bin encoding, 64-profile encoding, \textsc{q}ual\textsc{c}omp with rate 10 bits/read,
7th-degree polynomial regression, \textsc{q}ual\textsc{c}omp with rate 6 bits/read, and
finally, 5th-degree through 0-degree polynomial regression.

@@

Most of the compression methods did not vary greatly in terms of the
number of reads that were mapped \emph{only} by the compression
method; however, there is a sizable difference in the amount of reads
that are originally mapped, but unmapped by the compressed methods.
\textsc{q}ual\textsc{c}omp with rate 100 bits/read results in the fewest missing
original read alignments (159).  Increasing the regression model
polynomial degree results in a decreasing amount of reads that are
originally mapped, but unmapped by the regression model (40,931 to
1,134 reads for 0-degree and 7th-degree, respectively).  There is no
such trend for reads that are mapped only by the regression model.

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the number
  of reads that are kept or discarded after applying compression with
  respect to the original, uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
  \cline{2-12}
& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
\cline{2-12}
 & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
\cline{2-12}
& proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
 & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
\cline{2-12}
& proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
\end{tabular}
\end{table*}

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

@@

\section*{Acknowledgement}
We thank the 2014 Bioinformatics
Exchange for Students and Teachers (\textsc{best}) Summer School, 
funded by the offices of the Dean of The Graduate School at University
of Maryland and the Rektor of University of T\"{u}bingen, where this 
research was initiated.
